Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2022-01-26 fstats_pca_lewontin.tex
part1 revisions

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Data-driven methods such as Principal Component Analysis (PCA) are an important population genetic tool to understand this method. 
Here, I contrast PCA with a set of statistics motivated by trees ($F$-statistics). Here, I show that these two methods are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that just using the first few PCs provides good population structure is sparse, and only a few components contribute to most statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete and less biased descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a, gopalan2022}.
In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985}. However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{gopalan2022}.

Particularly for large and heterogeneous data sets, disentangling all these processes is challenging, and we cannot expect to devise a single model catching both broad strokes and minute details of human history. A commonly used analysis paradigm is thus to integrate tools based on different sets of assumptions. each emphasizing particular aspects of the data.

A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}. Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}). However, these methods are not designed to answer specific research questions, and are limited in their ability to estimate biologically meaningful parameters. For this purpose, methods based on explicit demographic models are often used that aim to fit a specified or estimated model of divergence, migration and genetic drift to the data \citep{gutenkunst2009, excoffier2013, kamm2015}. The drawback of these methods is that, to make inference mathematically feasible, we need to introduce strong modeling assumptions such as that populations are discrete, randomly mating, or at equilibrium. While in most cases these assumptions are violated to some extent and cannot be verified, we hope that the resulting model fits provide sufficiently accurate answers to specific research questions.

%\paragraph{F-statistics}
However, when the number of populations exceeds a few dozen,  even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the available data remains unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to build up complex models from smaller building blocks based on the relationship between two, three or four populations.

The framework is based on a set of parameters called  $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. Formal definitions will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, $F$-statistics are estimated from  data, and then used as tests of treeness. In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations. The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). However, admixture graphs are not the only plausible alternative models, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.

\paragraph{F-statistics and PCA}
The practical issue addressed in this study is how $F$-statistics can be reconciled with PCA, one of the most widely used data-driven modeling  techniques. One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal component, PC) retaining a maximum of the variance present in the data. To understand population structure, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize genetic diversity \citep{cavalli-sforza1994}. Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale data sets. 
%This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g. remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses. 
The PCA-decomposition has been studied for a number of explicit population genetic models including trees \citep{cavalli-sforza1975}, spatially continuous structure \citep{novembre2008a}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}. Here, in order to link PCA to $F$-statistics, I interpret both of them geometrically in \emph{allele frequency space}, i.e. as functions of a high-dimensional Euclidean space. For $F$-statistics, this interpretation was recently developed by \cite{oteo-garcia2021}, and for PCA it follows naturally from the interpretation of approximating a high-dimensional space with a low-dimensional one.

@@

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fstats_pca_vs_tree.pdf}
	\caption{\textbf{Representation of $F$-statistics on trees and 2D-PCA-plots.} The schematics show four populations and their representation using a tree (top row) or a 2D-PCA plot (bottom row). A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space. B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthogonal projection of $X_1 - X_3$ onto $X_1$-$X_4$. C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or the orthogonal projection of $X_2 - X_3$ on $X_1 - X4$. D: $F_4(X_1, X_2; X_3, X_4)$ The two paths from $X_1$ to $X_2$ and $X_3$ and $X_4$ are non-overlapping in the tree, which corresponds to orthogonal vectors in PCA-space.}
	\label{fig:geom}
\end{figure}

@@

\subsection{Formal Definition of $F$-statistics}
Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ loci. Let $x_{il}$ denote the frequency at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$. As $X_i$ will be the only data summary considered here for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.

@@

$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch, respectively (Figure \ref{fig:geom}C). Crucially, for branches that do not exist in the tree (as in Figure \ref{fig:geom}D), $F_4$ will be zero. The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}.

@@

The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A. Here, the (typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe. This will result in $F_4$-statistics that are non-zero, and, in some cases, in negative $F_3$-statistics \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Recently, \cite{oteo-garcia2021}  re-derived $F$-statistics in a geometric framework, showing that these assumptions are not necessary. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to the vector projection
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:geom}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. On trees, all disjoint paths are independent (i.e. orthogonal) from each other, and thus the external branches vanish under the projection.

The drawback of the geometric approach of \cite{oteo-garcia2021} is that we have to deal with an very high-dimensional space, as the number of SNPs is frequently in the millions. However, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data \citep{patterson2006}. Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g. in how SNPs are standardized, how missing data is treated, what algorithm is used, or whether we use individuals or populations as units of analysis. The version of PCA I use here is set up such that the similarities to $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies. In contrast, many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ of many populations in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each population $X_i$
of $\MX$ as a point in $\mathbb{R}^S$.

PCA allows us to approximate the points in the high-dimensional allele frequency space by a $K$-dimensional subspace of the  data. If all PCs are considered, $K=n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

\subsection{Connection between PCA and $F$-statsitics}	
\subsubsection{Principal components from $F$-statistics}
PCA, as defined above, and $F$-statistics are closely related. In fact, the principal components can be directly calculated from $F$-statistics using multidimensional scaling, which, for squared Euclidean ($F_2$)-distances, leads to an identical decomposition to PCA \citep{gower1966}. Suppose we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$. We can obtain the principal components from this matrix by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
\begin{equation}
\MP\MP^T = - \frac{1}{2}\MC\MF\MC \text{.} \label{eq:mds}
\end{equation}

@@

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) &=& \sum_{k=1}^K(p_{ik} - p_{jk})^2 &+& \sum_{k=K+1}^n(p_{ik} - p_{jk})^2&&&&\nonumber\\
&=& {\hat{F_2}^{(K)}(P_i, P_j)} &+& {\epsilon^{(K)}(P_i, P_j)}&&&& \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript of $\hat{F_2}$ when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations in our sample, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the sum of approximation errors of all $F_2$-statistics.

@@

\section{Material \& Methods}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation in PCA-space, which can be approximated on PCA plots. In the next section I explore this connection in detail, and illustrate it on two sample data sets that I briefly introduce here. Both are based on the analyses by \cite{lazaridis2014}. The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs). SNPs with missing data in any population are excluded. The code used to write this paper, create all figures and analyses is  available on \url{https://github.com/BenjaminPeter/fstats_pca}.

@@

\paragraph{Western Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe.  \cite{lazaridis2014} used this data set as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}. Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}. I thus speculate that gene flow between these populations is common \citep{ralph2013}, and a discrete model such as a tree or an admixture graph might be a rather poor reflection of this data. A file with all individuals used and their assigned population is given in \textbf{Supplementary File 2.}

>>>>

\title{Modelling complex population structure using $F$-statistics and Principal Component Analysis}
\author{Benjamin M Peter}
\begin{document}
	\maketitle
\begin{abstract}
Human  genetic diversity is shaped by our complex history. Population genetic tools to understand this variation can broadly be classified into data-driven methods such as Principal Component Analysis (PCA), and model-based approaches such as $F$-statistics \textit{sensu} Patterson. I show that these two perspectives are closely related, and I derive explicit connections between the two approaches. I show that $F$-statistics have a simple geometrical interpretation in the context of PCA, and that orthogonal projections are the key concept to establish this link. I illustrate my results on two examples, one of local, and one of global human diversity. In both examples, I find that the first few PCs are sufficient to approximate most $F$-statistics, and thus PCA-plots are effective to predict $F$-statistics. Based on these results, I develop novel visualizations that allow for investigating specific hypotheses, checking the assumptions of more sophisticated models. My results extend $F$-statistics to non-discrete populations, moving towards more complete descriptions of human genetic variation. 
\end{abstract}
\section{Introduction}
As in most species, the genetic diversity of human populations has been influenced by our history and environment over the last several hundred thousand years \citep[e.g][]{cavalli-sforza1994,marciniak2017, reich2018a, nielsen2017, witt2022}. In turn, an important goal of population genetics is to use observed patterns of variation to  investigate and reconstruct the demographic and evolutionary history of our species \citep{schraiber2015, orlando2021}.

The complicated genetic structure observed in present-day human populations \citep{the2015, mallick2016} is caused by the interplay of demographic and evolutionary processes with both discrete and continuous components \citep{pritchard2000, rosenberg2002a, serre2004, rosenberg2005, bradburd2018, reich2018a, peter2020a, gopalan2022}.
In particular, populations are expected to slowly differentiate if they are isolated from each other \citep{wahlund1928,cavalli-sforza1975}. In humans, this may be caused because continental-scale geographic distances limit migration, causing a pattern known as isolation-by-distance \citep{slatkin1985}. However, these patterns are usually not uniform, but shaped by geography, particularly barriers  to migration such as mountain ranges, oceans or deserts \citep{cavalli-sforza1994, barbujani1990, rosenberg2005, bradburd2013, peter2020a}. In addition,  major historical population movements such as the out-of-Africa, Austronesian or Bantu expansions lead to more gradual patterns of genetic diversity over space \citep{cavalli-sforza1994, ramachandran2005, novembre2008, stoneking2016, racimo2020}. Local migration between neighboring populations will reduce differentiation, and long-distance migrations \citep{alves2016}, and secondary contact between diverged populations, such as Neandertals and modern humans \citep{green2010} may lead to locally increased diversity \citep{boca2020}.

Particularly for large and heterogeneous data sets, disentangling all these processes is challenging, and we cannot expect to devise a single model catching both broad strokes and minute details of human history. A commonly used analysis paradigm is thus to combine tools based on different sets of assumptions, each emphasizing particular aspects of the data.

A typical analysis starts with data-driven, exploratory methods that summarize data making minimal assumptions \citep[e.g.][]{schraiber2015}. Examples are population trees \citep{cavalli-sforza1967, felsenstein1973, cavalli-sforza1975}, Principal Component Analysis \citep[PCA, ][]{cavalli-sforza1994, patterson2006}) structure-like models \citep{pritchard2000, alexander2009} or multidimensional scaling \citep[MDS][]{lessa1990}. These methods are limited in their ability to estimate biologically meaningful parameters, but provide useful summaries and visualizations. These analyses are then complemented with methods based on explicit demographic models, which are used to estimate parameters or test hypotheses \citep{gutenkunst2009, excoffier2013, kamm2015}.

\paragraph{F-statistics}
When the number of populations exceeds a few dozen, even codifying reasonable population models can be prohibitively difficult. One approach is to pick a small set of ``representative'' samples, and restrict modeling to this subset \citep[e.g.][]{gravel2011, harney2021}. However, this has the drawback that a large proportion of the available data remains unused. An increasingly popular alternative approach, particularly in the analysis of human ancient DNA, is therefore to focus on the relationship between two, three or four populations, commonly using $F$-statistics \textit{sensu} Patterson \citep{reich2009, patterson2012, peter2016}. Formal definition will be given in the Theory section; but an informal motivation starts with the null model that populations are related as a tree, in which each $F$-statistic measures the length of a particular set of branches. \citep[Figure \ref{fig:geom};][]{semple2003, peter2016}.

In most applications, $F$-statistics are estimated from data, and then used as tests of treeness. In particular, under the assumption of a tree, $F_3$ is restricted to be non-negative, and many $F_4$-statistics will be zero \citep{semple2003, patterson2012}, and data that violates these constraints is incompatible with a tree-like relationship between populations. The canonical alternative model is an admixture graph (or phylogenetic network) \citep{patterson2012, huson2010}, which is a tree which allows for additional edges reflecting gene flow (Figure \ref{fig:admix}A). However, admixture graphs are not the only plausible alternative model, and expected $F$-statistics can be calculated for a wide range of population genetic demographic models \citep{peter2016}.

\paragraph{F-statistics and PCA}
The practical issue addressed in this study is how $F$-statistics can be compared with PCA, one of the most widely used data-driven modeling  techniques. One way PCA can be motivated is as generating a low-dimensional representation of the data, with each dimension (called principal component, PC) retaining a maximum of the variance present in the data. To understand population structure, the use of PCA has been pioneered by \cite{cavalli-sforza1964}, who used allele-frequency data at a population level to visualize genetic diversity \citep{cavalli-sforza1994}. Currently, PCA is most commonly performed on individual-level genotype data \citep[e.g.][]{patterson2006, novembre2008}, making use of the hundreds of thousands of loci available in most genome-scale data sets. 
%This is most commonly done in early, exploratory phases of a study \citep{schraiber2015}, as PCA can be used to e.g. remove outliers that do not cluster with their population (due to technical or biological reasons), but it remains an important part of many analyses. 
The PCA-decomposition has been studied for a number of explicit population genetic models including trees \citep{cavalli-sforza1975}, spatially continuous structure \citep{novembre2008a}, the coalescent \citep{mcvean2009} and discrete population models \citep{francois2021}. Here, in order to link PCA to $F$-statistics, I interpret both of them geometrically in \emph{allele frequency space}, i.e. as functions of a high-dimensional Euclidean space. For $F$-statistics, this interpretation was recently developed by \cite{oteo-garcia2021}, and for PCA it follows naturally from the interpretation of approximating a high-dimensional space with a low-dimensional one.

@@

\begin{figure}[!ht]
	\includegraphics[width=\textwidth]{figures/fstats_pca_vs_tree.pdf}
	\caption{\textbf{Representation of $F$-statistics on trees and 2D-PCA-plots.} The schematics show four populations and their representation using an (arbitrarily rooted) tree (top row) or a 2D-PCA plot (bottom row). A: $F_2$ represents the (squared) Euclidean distance between two tree leafs, and in PC-space. B: $F_3(X_1; X_3, X_4)$ corresponds to the external branch from $X_1$ to the internal node joining the populations, and is proportional to the orthogonal projection of $X_1 - X_3$ onto $X_1$-$X_4$. C: $F_4(X_1, X_4; X_2, X_3)$ corresponds to the internal branch in the tree, or the orthogonal projection of $X_2 - X_3$ on $X_1 - X4$. D: $F_4(X_1, X_2; X_3, X_4)$ The two paths from $X_1$ to $X_2$ and $X_3$ and $X_4$ are non-overlapping in the tree, which corresponds to orthogonal vectors in PCA-space.}
	\label{fig:geom}
\end{figure}

@@

\subsection{Formal Definition of $F$-statistics}
Let us assume we have a set of populations for which we have SNP allele frequency data from $S$ biallelic loci. Let $x_{il}$ denote the frequency of an arbitrary allele at the $l$-th SNP in the $i$-th population; and let $X_i = (x_{i1}, x_{i2}, \dots x_{iS})$  be a vector collecting all allele frequencies for population $i$. As $X_i$ will be the only data summary considered here for population $i$, I make no distinction between the population and the allele frequency vector used to represent it.

@@

$F$-statistics have been primarily motivated in the context of trees and admixture graphs \citep{patterson2012}. In a tree, the squared Euclidean distance $F_2(X_1, X_2)$ measures the length of the path between populations $X_1$ and $X_2$ (Figure \ref{fig:geom}A); $F_3$ represents the length of an external branch (Figure \ref{fig:geom}B) and $F_4$ the length of an internal branch, respectively (Figure \ref{fig:geom}C). The length of each branch can be thought of in units of genetic drift, and is non-negative \citep{patterson2012}. Crucially, this means that $F_4$ will be zero for pairs of populations from non-overlapping clades, which means that the tree lacks the branch corresponding to this statistic (as in Figure \ref{fig:geom}D).

@@

The most widely considered alternative model is an admixture graph \citep{patterson2012}, an example is given in Figure \ref{fig:admix}A. Here, the *typically unobserved) population $X_y$ is generated by a mixture of individuals from the ancestors of $X_2$ and $X_3$. Over time, genetic drift will change $X_y$ to $X_x$, which is the admixed population we observe. In this case, all $F_4$-statistics involving $X_y$ and both admixture sources will be non-zero, and, in some cases, $F_3(X_y; X_2, X_3)$ will be negative \citep[exact conditions can be found in ][]{peter2016}.

\subsubsection{Geometric interpretation of $F$-statistics}
An implicit assumption in the development of $F$-statistics is that population lineages are mostly discrete, and that gene flow is rare. Specifically, they interpret the populations $X_i$ as points or vectors in the $S$-dimensional \emph{allele frequency space} $\mathbb{R}^S$. In this case, the $F$-statistics can be thought of as inner (or dot) products, and they showed that all properties and tests related to treeness can be derived in this larger space. In particular the $F$-statistics can be written as
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
	\begin{align}
	\langle a, b \rangle &= \sum_i a_ib_i\\
	\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
	\langle a, a \rangle &= \normsq{a}\\
	\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
	\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$. The inner product is  closely related to vector projections
\begin{equation}
\vectorproj[b]{a} = \frac{\langle a , b\rangle}{\normsq{b}} b,\label{eq:proj}
\end{equation}
which is a vector colinear to $b$ whose length measures how much vector $a$ points in the direction of $b$. Thinking of $F$-statistics as projections also holds on trees: In e.g. a $F_4(X_1, X_4; X_2, X_3)$-statistic (Figure \ref{fig:admix}C), the internal branch is precisely the intersection of the paths from $X_1$ to $X_4$ and from $X_2$ and $X_3$. The external branches are independent populations, and so they are expected to drift orthogonally to each other.

The geometric approach of \cite{oteo-garcia2021} assumes a very high-dimensional space, as each SNP in a data set adds a dimension, and data sets have commonly a million or more SNPs. Such high-dimensional spaces are hard to visualize and analyze, however, it has been commonly observed that population structure is quite low-dimensional, and that the first few PCs provide a good approximation of the covariance structure in the data \citep{patterson2006}. Therefore, we may hope that PCA could yield a reasonable approximation of the allele frequency space, and that $F$-statistics as measures of population structure may likewise be well-approximated by the first few PCs.

\subsection{Formal Definition of PCA}
PCA is a common way of summarizing genetic data, and so a large number of variations of PCA exist, e.g. in how SNPs are standardized, how missing data is treated or whether we use individuals or populations as units of analysis \citep{cavalli-sforza1994, patterson2006}. The version of PCA I use here is set up such that the similarities to  $F$-statistics are maximized, and does \emph{not} reflect how PCA is most commonly applied to genome-scale human genetic variation data sets. In particular, I assume that a PCA is performed on unscaled, estimated population allele frequencies, whereas many applications of PCA are based on individual-level sample allele frequency, scaled by the estimated standard deviation of each SNP \citep{patterson2006}. The differences this causes will be addressed in the discussion.

Let us again assume we have allele frequency data as above, but let us now assume we aggregate the allele frequency vectors $X_i$ of many populations in a matrix $\MX$ whose entry $x_{il}$ reflects the allele frequency of the $i$-th population at the $l$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. Since the allele frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}, which is a subset of $\mathbb{R}^S$.

One way PCA can be motivated is that it aims to find a $K$-dimensional subspace of the data space that retains most  variation in the data. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{low-rank} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

\subsection{Connection between PCA and $F$-statsitics}	
\subsubsection{Principal components from $F$-statistics}
PCA, as defined above, and $F$-statistics are closely related. It is a classical result that PCA is equivalent to multidimensional scaling using squared Euclidean distances \citep{gower1966}. Since $F_2$-distances are squared Euclidean, we calculate the pairwise $F_2(X_i, X_j)$ between all $n$ populations, and collect them in a matrix $\MF$. Multidimensional scaling then proceeds by double-centering it, so that its row and column means are zero, and perform an eigendecomposition of the resulting matrix:
\begin{equation}
\MP\MP^T = - \frac{1}{2}\MC\MF\MC \text{.} \label{eq:mds}
\end{equation}
Although we arrive at $\MP$ from a very different angle, as long as we make the same choices about normalization and units of analysis, we will get the exact same results.

@@

In most applications, we do not use all PCs, but instead truncate to the first $K$ PCs, which explain most of the between-population genetic variation.
Thus, 
\begin{align}
F_2(P_i, P_j) = \underbrace{\sum_{k=1}^K(p_{ik} - p_{jk})^2}_{\hat{F_2}^{(K)}(P_i, P_j)} + \underbrace{\sum_{k=K+1}^n(p_{ik} - p_{jk})^2}_{\epsilon^{(K)}(P_i, P_j)} \text{.}
\end{align}
In this notation, $\hat{F_2}^{(K)}$ is the approximation of $F_2$ with only the first $K$ PCs considered, and $\epsilon^{(K)}$ is the corresponding approximation error. I will omit the superscript of $\hat{F_2}$ when the exact number of PCs is not relevant.
If we sum up the squared approximation errors over all pairs of populations in our sample, we obtain 
\begin{equation}
\sum_{i,j} \epsilon^{(K)}(P_i, P_j)^2 = \sum_{i,j} \left(\hat{F_2}^{(K)}(P_i, P_j) - F_2^{(K)}(P_i, P_j)\right)^2 = \normsq{\MF - \hat{\MF}}_F \text{,}
\end{equation}
where the Frobenius-norm $\normsq{\cdot}_F$ of a matrix is defined as the square root of the sum-of-squares of all its elements. This is precisely the function that is minimized in MDS \citep{jolliffe2013}. In that sense, $\hat\MF^{(K)}$ is the optimal low-rank approximation of $\MF$ for any $K$ in that it minimizes the sum of approximation errors of all $F_2$-statistics.

@@

\section{Material \& Methods}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation in PCA-space, which can be approximated on PCA plots. In the next section I explore this connection in detail, and illustrate it on two sample data sets that I briefly introduce here. Both are based on the analyses by \cite{lazaridis2014}. The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs). SNPs with missing data in any population are excluded. The code used to create all figures and analyses will be available on \url{https://github.com/BenjaminPeter/fstats_pca}.

@@

\paragraph{Western Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe. It is motivated by the analysis of \cite{lazaridis2014}, who used it as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}. Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}. I thus speculate that gene flow between these populations is common \citep{ralph2013}, and a discrete model such as a tree or an admixture graph might be a rather poor reflection of this data. A file with all individuals used and their assigned population is given in \textbf{Supplementary File 2.}
